---
title: "Chapter 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
```

## First steps

1. Run ggplot(data = mpg). What do you see?

```{r first-steps-1}
ggplot(data = mpg)
```

We see nothing because no aesthetics have been added.


2. How many rows are in mpg? How many columns?

```{r first-steps-2}
nrow(mpg)
ncol(mpg)
```


3. What does the drv variable describe? Read the help for ?mpg to find out.

"the type of drive train, where f = front-wheel drive, r = rear wheel drive, 4 = 4wd"


4. Make a scatterplot of hwy vs cyl.

```{r first-steps-4}
ggplot(data = mpg) +
    geom_point(aes(x = hwy, y = cyl))
```


5. What happens if you make a scatterplot of class vs drv? Why is the plot not useful?

```{r first-steps-5}
ggplot(data = mpg) +
    geom_point(aes(x = class, y = drv))
```

This graph is not useful because points are overlapping and there is also no
discernible pattern in the data with this plot.

## Aesthetic mappings

1. What’s gone wrong with this code? Why are the points not blue?

```{r aesthetic-mappings-1a}
ggplot(data = mpg) + 
    geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```

The position of the parentheses has a mistake. Correct is:

```{r aesthetic-mappings-1b}
ggplot(data = mpg) + 
    geom_point(aes(x = displ, y = hwy), color = "blue")
```


2. Which variables in mpg are categorical? Which variables are continuous? (Hint: type ?mpg to read the documentation for the dataset). How can you see this information when you run mpg?

- manufacturer
- model
- trans
- drv
- fl
- class

```{r aesthetic-mappings-2}
mpg
```

These are character variables (*\<chr\>*).


3. Map a continuous variable to color, size, and shape. How do these aesthetics behave differently for categorical vs. continuous variables?

```{r aesthetic-mappings-3a}
ggplot(data = mpg) + 
    geom_point(aes(x = cty, y = hwy, color = displ))

ggplot(data = mpg) + 
    geom_point(aes(x = cty, y = hwy, size = displ))
```

Mapping to shape of a continuous variable gives an error:
"A continuous variable can not be mapped to shape"

```{r aesthetic-mappings-3b}
ggplot(data = mpg) + 
    geom_point(aes(x = cty, y = hwy, color = fl))

ggplot(data = mpg) + 
    geom_point(aes(x = cty, y = hwy, size = fl))

ggplot(data = mpg) + 
    geom_point(aes(x = cty, y = hwy, shape = fl))
```

Although it is possible to map a categorical variable to size, it does not make
much sense, at least in this case.

Moreover, in the case of a continuous variable mapping to color also give a 
continuous color scale. Mapping a categorical variable to color give as many
colors as there are categories in that variable.

4. What happens if you map the same variable to multiple aesthetics?

```{r aesthetic-mappings-4}
ggplot(data = mpg) + 
    geom_point(aes(x = cty, y = hwy, color = fl, size = fl, shape = fl))
```

The different aesthetics are combined then.

5. What does the stroke aesthetic do? What shapes does it work with? (Hint: use ?geom_point)

```{r aesthetic-mappings-5}
ggplot(data = mpg) + 
    geom_point(aes(x = cty, y = hwy, color = fl, stroke = cyl), fill = "black", shape = 21)
```

Stroke works with a few shapes of the point geometry. For these specific shapes
it determines the width of the border of the shape.

6. What happens if you map an aesthetic to something other than a variable name, like aes(colour = displ < 5)? Note, you’ll also need to specify x and y.

```{r aesthetic-mappings-6}
ggplot(data = mpg) + 
    geom_point(aes(x = cty, y = hwy, color = displ < 5))
```

Then the aesthetic (color here) is determined by whether the condition is true or false.
